---
title: "R Notebook of Concrete regression"
author: "David Stelzig"
output: html_notebook
---

# Introduction

This is an R Notebook that demonstrates the basic analysis and regression of the concrete dataset: 
Yeh, I. (1998). Concrete Compressive Strength [Dataset]. UCI Machine Learning Repository. https://doi.org/10.24432/C5PK67.

## Dataset

The data contain 8 characteristic, numeric features contributing to the target variable *concrete compressive strength*, including cement, water, Superplasticizer and more.

## Topics

The notebook includes the following steps:

-   Import of libraries, loading the data and **cleaning** the data (missing values, useless columns, duplicates ...)
-   **Exploratory Data Analysis**
-   **Training** of a different regression models and multi-component analysis

# 1. Load Libraries and import the data
```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(e1071)
library(ggplot2)
library(corrplot)
library(gbm)
library(class)
library(tidyr)
library(dplyr)
library(readxl)
library(GGally)
concrete_data <- read_excel("C:/Users/Sabrina/OneDrive/Desktop/David_work/R/Concrete_Data.xls")

head(concrete_data)
```
Rename the columns for better readability:

```{r}
colnames(concrete_data) <- c("Cement", "BlastFurnaceSlag", "FlyAsh", "Water", "Superplasticizer", 
                             "CoarseAggregate", "FineAggregate", "Age", "Strength")
head(concrete_data)
```
# 2. Inspect the data, data Cleaning and Preprocessing
```{r}
str(concrete_data)
```
The dataset contain 1030 entries with only numerical values.
```{r}
summary(concrete_data)
```
Check for missing values
```{r}
colSums(is.na(concrete_data))
```
The dataset does **not** contain missing values.
Check for duplicates and remove them:
```{r}
duplicates <- concrete_data[duplicated(concrete_data), ]
concrete_data <- concrete_data %>% distinct()
str(concrete_data)
```
The dataset contained **25 duplicates**. After dropping them 1005 entries remain.

# 3. Exploratory Data Analysis

Check the overall correlation and distribution of the data
```{r}
library(ggplot2)
ggpairs(concrete_data,
        columns = 1:9,
        lower = list(continuous = wrap("points", alpha = 0.5, size = 1.8, color = "#030e45")),
        upper = list(continuous = wrap("cor", size = 4, color = "#648fff", display_grid = FALSE)),
        axisLabels = "show") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 6, face = "bold"),
        axis.text = element_text(size = 6))
```
At first glance it is easy to extract the following correlation:
- positive correlation between **Cement** and **Strength**
- positive correlation between **Strength** and **Age**
- positive correlation between **Strength** and **Superplasticizer**
- Negative correlation between **Water** and **Superplasticizer**
- Negative correlation between **Water** and **FineAggregate**

Compute a correlation matrix for better visualization:
```{r}
cor_matrix <- cor(concrete_data)
corrplot::corrplot(cor_matrix,
                   method = "color",
                   type = "upper",
                   tl.cex = 0.8,
                   number.cex = 0.7,
                   addCoef.col = "white",
                   tl.col = "#240046",
                   number.digits = 2,
                   col = colorRampPalette(c("#240046", "#3c096c", "#5a189a","#7b2cbf","#9d4edd","#c77dff"))(200),
                   bg = "white",
                   outline = "#240046",
                   tl.srt = 45,
                   cl.cex = 0.8,
                   cl.align.text = "l")
```
Check the correlation between **Age** and **Strength**:
```{r}
ggplot(concrete_data, aes(x = Age, y = Strength)) +
  geom_point(alpha = 0.7, size = 2.5, color = "#00b4d8") +  # Enhanced points
  geom_smooth(method = "loess", color = "#030e45", fill = "#648fff", alpha = 0.4, linewidth = 1.2) + # Enhanced smooth line
  labs(title = "Concrete Strength vs. Age",
       x = "Age (days)",
       y = "Strength (MPa)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.major = element_line(color = "#e0e0e0", linewidth = 0.2),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white"))
```
Visually it looks like the relationship is not perfectly linear.

# 4.Machine Learning Models

## Question 1: Can we predict the strength based on other variables? -> **Regression**
```{r}
# Split data into training and testing sets
set.seed(123)  # For reproducibility
train_indices <- sample(seq_len(nrow(concrete_data)), size = 0.8 * nrow(concrete_data))
train_data <- concrete_data[train_indices, ]
test_data <- concrete_data[-train_indices, ]
```
### Linear Regression
```{r}
linear_model <- lm(Strength ~ ., data = train_data)
summary(linear_model)

linear_predictions <- predict(linear_model, newdata = test_data)
print(paste("Linear Regression RMSE:", RMSE(linear_predictions, test_data$Strength)))
r2_lr_model <- cor(test_data$Strength, linear_predictions)^2
print(paste("Linear Regression RMSE R²:", round(r2_lr_model, 3)))
```
The model performs mediocre with an R2 of 0.595.
Create visualization to check the deviations between actual data and the fit.
```{r}
ggplot(test_data, aes(x = Strength, y = linear_predictions)) +
  geom_point(alpha = 0.7, size = 2.5, color = "#00b4d8") +  # Enhanced points
  geom_abline(intercept = 0, slope = 1, color = "#030e45", linewidth = 1.2, linetype = "dashed") + # Enhanced abline
  labs(title = "Actual vs. Predicted Strength",
       x = "Actual Strength (MPa)",
       y = "Predicted Strength (MPa)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        panel.grid.major = element_line(color = "#e0e0e0", linewidth = 0.2),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white"))
```
```{r}
# Compute residuals
residuals <- test_data$Strength - predict(linear_model, newdata = test_data)

# Create residual plot
ggplot(data.frame(Predicted = predict(linear_model, newdata = test_data), Residuals = residuals), 
       aes(x = Predicted, y = Residuals)) +
  geom_point(alpha = 0.7, color = "#7209b7", size = 2.5) +
  geom_hline(yintercept = 0, color = "#030e45", linetype = "dashed",linewidth = 1.2,) +
  labs(title = "Residual Plot", x = "Predicted Strength (MPa)", y = "Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
```
It is apparent that the model doesn't perform perfectly.
Try other.

### Random Forest Regression
```{r}
rf_model_reg <- randomForest(Strength ~ ., data = train_data)
rf_predictions_reg <- predict(rf_model_reg, newdata = test_data)
print(paste("Random Forest RMSE:", RMSE(rf_predictions_reg, test_data$Strength)))
r2_rf_model <- cor(test_data$Strength, rf_predictions_reg)^2
print(paste("Random Forest R²:", round(r2_rf_model, 3)))
```
Random Forest performs much better with an R2 of 0.916.
```{r}
ggplot(test_data, aes(x = Strength, y = rf_predictions_reg)) +
  geom_point(alpha = 0.7, size = 2.5, color = "#52b788") +  # Enhanced points
  geom_abline(intercept = 0, slope = 1, color = "#1b4332", linewidth = 1.2, linetype = "dashed") + # Enhanced abline
  labs(title = "Actual vs. Predicted Strength",
       x = "Actual Strength (MPa)",
       y = "Predicted Strength (MPa)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        panel.grid.major = element_line(color = "#e0e0e0", linewidth = 0.2),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white"))
```
```{r}
residuals_rf <- test_data$Strength - predict(rf_model_reg, newdata = test_data)

# Create residual plot
ggplot(data.frame(Predicted = predict(rf_model_reg, newdata = test_data), Residuals = residuals_rf), 
       aes(x = Predicted, y = Residuals)) +
  geom_point(alpha = 0.7, color = "#52b788", size = 2.5) +
  geom_hline(yintercept = 0, color = "#1b4332", linetype = "dashed",linewidth = 1.2,) +
  labs(title = "Residual Plot", x = "Predicted Strength (MPa)", y = "Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))

```
###Gradient Boosting Regression
```{r}
gbm_model_reg <- gbm(Strength ~ ., data = train_data, distribution = "gaussian", n.trees = 100, interaction.depth = 3, shrinkage = 0.1)
gbm_predictions_reg <- predict(gbm_model_reg, newdata = test_data, n.trees = 100)
print(paste("GBM RMSE:", RMSE(gbm_predictions_reg, test_data$Strength)))
r2_gbm_model <- cor(test_data$Strength, gbm_predictions_reg)^2
print(paste("Random Forest R²:", round(r2_gbm_model, 3)))
```

```{r}
ggplot(test_data, aes(x = Strength, y = gbm_predictions_reg)) +
  geom_point(alpha = 0.7, size = 2.5, color = "#ad2831") +  # Enhanced points
  geom_abline(intercept = 0, slope = 1, color = "#38040e", linewidth = 1.2, linetype = "dashed") + # Enhanced abline
  labs(title = "Actual vs. Predicted Strength",
       x = "Actual Strength (MPa)",
       y = "Predicted Strength (MPa)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        panel.grid.major = element_line(color = "#e0e0e0", linewidth = 0.2),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white"))
```
Given the three tested models, **Random Forest** performed best and will be used further. 
However, for other data it might be wise to test also other regression models.

### Feature Importance Analysis
```{r}
importance(rf_model_reg)
varImpPlot(rf_model_reg)
```
**Age**, **Cement** and **Water** have the most significant effect on the outcome of the model.

### Model Optimization

```{r}
# Tuning Random Forest Hyperparameters
set.seed(123)
best_mtry <- tuneRF(train_data[, -9], train_data$Strength, stepFactor = 1.5, improve = 0.01, trace = TRUE)

set.seed(123)
rf_model <- randomForest(Strength ~ ., 
                         data = train_data, 
                         mtry = 4, 
                         ntree = 2000,  
                         nodesize = 3,  
                         importance = TRUE)  
```
Given the **tuneRF** function and the output of a minimum mtry value of 4, the randomForest model will be trained with the optimized parameters.

```{r}
# Predictions on the test set
predictions_rf <- predict(rf_model, newdata = test_data)

# Compute R² and RMSE
r2_rf <- cor(test_data$Strength, predictions_rf)^2
rmse_rf <- sqrt(mean((predictions_rf - test_data$Strength)^2))


print(paste("Optimized Random Forest R²:", round(r2_rf, 3)))
print(paste("Optimized Random Forest RMSE:", round(rmse_rf, 3)))
print(paste("Random Forest RMSE:", RMSE(rf_predictions_reg, test_data$Strength)))
print(paste("Random Forest R²:", round(r2_rf_model, 3)))
```
**Hyperparameter Tuning** has helped to improve the accuracy of the model.
```{r}
ggplot(test_data, aes(x = Strength, y = predictions_rf)) +
  geom_point(alpha = 0.7, size = 2.5, color = "#52b788") +
  geom_abline(intercept = 0, slope = 1, color = "#1b4332", linewidth = 1.2, linetype = "dashed") +
  labs(title = "Actual vs. Predicted Strength",
       x = "Actual Strength (MPa)",
       y = "Predicted Strength (MPa)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        panel.grid.major = element_line(color = "#e0e0e0", linewidth = 0.2),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white"))
```

## Question 2:  How do different material compositions (e.g., Cement vs. Fly Ash vs. Blast Furnace Slag) interact to influence strength? 
```{r}
rf_model <- randomForest(Strength ~ ., data = train_data, importance = TRUE)
importance(rf_model)
varImpPlot(rf_model)
```
```{r}
library(pdp)
# Partial dependence of Cement
cement_pdp <- partial(rf_model, pred.var = "Cement")
plot(cement_pdp)
```
**Increasing** the Cement content generally leads to increased *Strength*.
```{r}
#Partial dependence of Water
Water_pdp <- partial(rf_model, pred.var = "Water")
plot(Water_pdp)
```
**Water content** should be reduced.

Check for partial dependence of water and cement.

```{r}
interaction_pdp <- partial(rf_model, pred.var = c("Cement", "Water"))
ggplot(interaction_pdp, aes(x = Cement, y = Water, z = yhat)) +
  geom_tile(aes(fill = yhat), color = "white") +
  scale_fill_viridis_c(option = "C") +
  labs(title = "Interaction Between Cement and Water on Compressive Strength",
       subtitle = "Partial Dependence Plot (PDP)",
       x = "Cement Content",
       y = "Water Content",
       fill = "Predicted Strength") +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  theme(legend.position = "right")

```
Check for partial dependence of Age and cement.


```{r}
interaction_pdp <- partial(rf_model, pred.var = c("Cement", "Age"))
ggplot(interaction_pdp, aes(x = Cement, y = Age, z = yhat)) +
  geom_tile(aes(fill = yhat), color = "white") +
  scale_fill_viridis_c(option = "viridis") +
  labs(title = "Interaction Between Cement and Age on Compressive Strength",
       subtitle = "Partial Dependence Plot (PDP)",
       x = "Cement Content",
       y = "Age",
       fill = "Predicted Strength") +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  theme(legend.position = "right")
```
## Question 3: How much does a Suplerplasticizer help? At what point does adding Superplasticizer not help anymore?

```{r}
superplasticizer_pdp <- partial(rf_model, pred.var = "Superplasticizer")
plot(superplasticizer_pdp)
range(concrete_data$Superplasticizer)

min_strength <- superplasticizer_pdp$yhat[which.min(superplasticizer_pdp$Superplasticizer)]
max_strength <- superplasticizer_pdp$yhat[which.max(superplasticizer_pdp$Superplasticizer)]

# Calculate the difference
strength_increase <- max_strength - min_strength
print(paste("Strength increase due to superplasticizer:", strength_increase))
```
```{r}
interaction_pdp <- partial(rf_model, pred.var = c("Water", "Superplasticizer"))
ggplot(interaction_pdp, aes(x = Water, y = Superplasticizer, z = yhat)) +
  geom_tile(aes(fill = yhat), color = "white") +
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "Interaction Between Cement and Age on Compressive Strength",
       subtitle = "Partial Dependence Plot (PDP)",
       x = "Water Content",
       y = "Superplasticizer",
       fill = "Predicted Strength") +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  theme(legend.position = "right")
```
Given the visualizations from above the water content can be kept minimum when using a superplasticizer. However, the use of the superplasticizer can be limited to ~ 15(component 5)(kg in a m^3 mixture) to achieve a high strength.

